Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV statistical mechanics

### 2019-11-13 paper2_JNLS/paper.tex
NM

% Dimensionless parameters
\newcommand{\ampscale}{\mathcal{A}}
\newcommand{\lengthscale}{\mathcal{L}}
\newcommand{\timescale}{\mathcal{T}}
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
% Hamiltonian stuff
\newcommand{\uhat}{\hat{u}}
\newcommand{\sympJ}{\mathcal{J}}
\newcommand{\vard}[2]{\frac{\delta #1}{\delta #2}}
\newcommand{\Ham}{\mathcal{H}}
\newcommand{\Hup}{\Ham^{-}}
\newcommand{\Hdn}{\Ham^{+}}
\newcommand{\Hupdn}{\Ham^{\pm}}
\newcommand{\Hthree}{\Ham_{3}}
\newcommand{\Htwo}{\Ham_{2}}
\newcommand{\Fcnl}{\mathcal{F}}
% Truncated stuff
\newcommand{\Proj}{\mathcal{P}_{\Lambda}}
\newcommand{\uL}{u_{\Lambda}}
\newcommand{\HLupdn}{\Ham_{\Lambda}^{\pm}}
\newcommand{\SympL}{\sympJ_{\Lambda}}
\newcommand{\uup}{u_{-}}
\newcommand{\udn}{u_{+}}
\newcommand{\uupdn}{u_{\pm}}
% Gibbs and theta
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gup}{\Gibbs^{-}}
\newcommand{\Gdn}{\Gibbs^{+}}
\newcommand{\Gupdn}{\Gibbs^{\pm}}
\newcommand{\thup}{\theta^{-}}
\newcommand{\thdn}{\theta^{+}}
\newcommand{\thupdn}{\theta^{\pm}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}
\newcommand{\meanupdn}[1]{\mean{#1}_{\pm}}
\newcommand{\Gz}{\Gibbs_0}
\newcommand{\meanz}[1]{\mean{#1}_0}
% Misc stuff
\newcommand{\tavg}[1]{\overline{#1}}
\newcommand{\transf}{\mathcal{F}}
\newcommand{\Nsamp}{N_s}
\newcommand{\sumsamp}{\sum_{i=1}^{\Nsamp}}
\newcommand{\traveldist}{\ell}
\newcommand{\Ftrans}{\mathcal{F}}
\newcommand{\RHS}{F}
\newcommand{\RHSh}{\hat{\RHS}}
\newcommand{\usqdah}{ (\widehat{u_{da}^2}) }

@@

Rearranging the TKdV system \eqref{TKdV} gives
\begin{align}
\label{TKdV2}
&\pd{\uL}{t} = -  \frac{1}{2} C_3 \drat^{-3/2} \, \pd{}{x} \Proj (\uL)^2 - C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = \RHS
%\qquad \text{for } x \in [-\pi,\pi]
\end{align}	
where we have represented the right-hand side by $\RHS$.
Our numerical scheme first separates the Hamiltonian into linear and nonlinear components. Then a pseudo-spectral scheme is used for spatial discretization with de-aliasing applied to the nonlinear term $\Proj (\uL)^2$ according to the standard 2/3-rule. That is, we first pad $\uhat_k$ with zeros for $\Lambda < \abs{k} \le 3\Lambda/2$, transform to physical space and square to obtain $\uL^2$ on a fine grid, then transform back to spectral space and truncate to obtain the Fourier coefficients of $\Proj (\uL)^2$
For simplicity, we denote these Fourier coefficients $\usqdah_k$. That is,
\begin{equation}
\Proj (\uL)^2 = \sum_{\abs{k} \le \Lambda} \usqdah_k e^{i k x}
\end{equation}
With this spectral discretization, the ODE system \eqref{TKdV2} can be written for the spectral vector $\mathbf{u}=\left\{ \hat{u}_{k}\right\}_{k=0}^{\Lambda}$ and $\mathbf{F}=\left\{ \RHSh_{k} \right\}_{k=0}^{\Lambda}$ as
\begin{equation}
\label{TKdVDisc}
\td{}{t} \uhat_k =  -\frac{1}{2} C_3 \drat^{-3/2} \, ik \usqdah_k + C_2 \drat^{1/2} \, ik^{3} \uhat_k = \RHSh_k
\end{equation}
Due to the nonlinearity, this system mixes scales during evolution. We note the third-order linear term may become stiff for large $\Lambda$.

For time integration of \eqref{TKdVDisc}, we employ a 4th-order midpoint symplectic scheme \cite{mclachlan1993symplectic}. Let $\mathbf{u}^{n}$ be the solution at time $t_{n}$ and $\mathbf{u}^{n+1}$
be the solution at next time step $t_{n+1}=t_{n}+\Delta t$, and $\mathbf{F}$ the
discretized operator for the spectral modes. The midpoint method has
two intermediate stages and the auxiliary vectors $\mathbf{y}_{1},\mathbf{y}_{2}$ so that
\begin{align}
&\mathbf{y}_{1}-\mathbf{u}^{n} = w_{1}\Delta t\mathbf{F}\left(\frac{1}{2}\left[\mathbf{y}_{1}+\mathbf{u}^{n}\right]\right),\\
&\mathbf{y}_{2}-\mathbf{y}_{1} = w_{2}\Delta t\mathbf{F}\left(\frac{1}{2}\left[\mathbf{y}_{2}+\mathbf{y}_{1}\right]\right),\\
&\mathbf{u}^{n+1}-\mathbf{y}_{2} = w_{3}\Delta t\mathbf{F}\left(\frac{1}{2}\left[\mathbf{u}^{n+1}+\mathbf{y}_{2}\right]\right),
\end{align}
with the time increments $w_{1}=\left(2+2^{1/3}+2^{-1/3}\right)/3$,
$w_{2}=1-2w_{1}$, and $w_{3}=w_{1}$.

@@

Then the PDFs can be counted from the solution $u_{\Lambda}\left(t\right)$, and the statistical expectations can be calculated by averaging along the trajectory
\begin{equation}
\left\langle F\left(u_{\Lambda}\right)\right\rangle _{\mu}=\frac{1}{T}\int_{t_{0}}^{t_{0}+T}F\left(u_{\Lambda}\left(t\right)\right)dt,
\end{equation}
given ergodicity in the solution with $u_{\Lambda}\left(t_{0}\right)$ already reaching the stationary state.

For the symplectic schemes, implicit iteration method is required to get the accurate solution. For any implicit problem of the form
\begin{equation}
\mathbf{z}-\mathbf{u}^{n}=\tau\mathbf{F\left(z\right)}=\tau\mathbf{Lz}+\tau\mathbf{N\left(z\right)},
\end{equation}
we split the operators into a linear and nonlinear components and only treat the nonlinear part on the right hand side for the previous stage in the iteration. Then we have the iterative scheme for the solution of implicit problem 
\begin{equation}
\left(1-\tau\mathbf{L}\right)\mathbf{z}^{\left[l+1\right]}=\mathbf{u}^{n}+\tau\mathbf{N}\left(\mathbf{z}^{\left[l\right]}\right).
\end{equation}
The linear operator $\mathbf{L}$ is the diagonal differential operator, thus it is easy to invert in these schemes. The termination criteria is determined by
\begin{equation}
\left\Vert \mathbf{z}^{\left[l+1\right]}-\mathbf{z}^{\left[l\right]}\right\Vert <\delta\;\mathrm{and}\;\left\Vert \mathbf{z}^{\left[l+1\right]}-\mathbf{z}^{\left[l\right]}\right\Vert \geq\left\Vert \mathbf{z}^{\left[l\right]}-\mathbf{z}^{\left[l-1\right]}\right\Vert .
\end{equation}
It is found that the iteration tolerance can be set as $\delta=1\times10^{-10}$.
Finally, the starting guess for the iterations can be found by quadratic extrapolation for the previous three stages $\left(t_{u},\mathbf{u}^{n}\right), \left(t_{u}-\tau,\mathbf{u}^{n-1}\right), \left(t_{u}-2\tau,\mathbf{u}^{n-2}\right)$. Therefore we can find the extrapolated value at $t=t_{n}+\beta\tau$ by
\begin{equation}
\mathbf{u}^{n+\beta}=\mathbf{Q}\left(t_{n}+\beta\tau\right)=\left[\frac{1}{2}\beta\left(\beta+3\right)+1\right]\mathbf{u}^{n}-\beta\left(\beta+2\right)\mathbf{u}^{n-1}+\frac{1}{2}\beta\left(\beta+1\right)\mathbf{u}^{n-2}.
\end{equation}

>>>>

% Dimensionless parameters
\newcommand{\ampscale}{\mathcal{A}}
\newcommand{\lengthscale}{\mathcal{L}}
\newcommand{\timescale}{\mathcal{T}}
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
% Hamiltonian stuff
\newcommand{\uhat}{\hat{u}}
\newcommand{\sympJ}{\mathcal{J}}
\newcommand{\vard}[2]{\frac{\delta #1}{\delta #2}}
\newcommand{\Ham}{\mathcal{H}}
\newcommand{\Hup}{\Ham^{-}}
\newcommand{\Hdn}{\Ham^{+}}
\newcommand{\Hupdn}{\Ham^{\pm}}
\newcommand{\Hthree}{\Ham_{3}}
\newcommand{\Htwo}{\Ham_{2}}
\newcommand{\Fcnl}{\mathcal{F}}
% Truncated stuff
\newcommand{\Proj}{\mathcal{P}_{\Lambda}}
\newcommand{\uL}{u_{\Lambda}}
\newcommand{\HLupdn}{\Ham_{\Lambda}^{\pm}}
\newcommand{\SympL}{\sympJ_{\Lambda}}
\newcommand{\uup}{u_{-}}
\newcommand{\udn}{u_{+}}
\newcommand{\uupdn}{u_{\pm}}
% Gibbs and theta
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gup}{\Gibbs^{-}}
\newcommand{\Gdn}{\Gibbs^{+}}
\newcommand{\Gupdn}{\Gibbs^{\pm}}
\newcommand{\thup}{\theta^{-}}
\newcommand{\thdn}{\theta^{+}}
\newcommand{\thupdn}{\theta^{\pm}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}
\newcommand{\meanupdn}[1]{\mean{#1}_{\pm}}
\newcommand{\Gz}{\Gibbs_0}
\newcommand{\meanz}[1]{\mean{#1}_0}
% Misc stuff
\newcommand{\tavg}[1]{\overline{#1}}
\newcommand{\transf}{\mathcal{F}}
\newcommand{\Nsamp}{N_s}
\newcommand{\sumsamp}{\sum_{i=1}^{\Nsamp}}
\newcommand{\traveldist}{\ell}
\newcommand{\Ftrans}{\mathcal{F}}
\newcommand{\RHS}{F}
\newcommand{\RHSh}{\hat{\RHS}}
\newcommand{\usqdah}{\hat{v}} %{ (\widehat{u_{da}^2}) }

@@

Rearranging the TKdV system \eqref{TKdV} gives
\begin{align}
\label{TKdV2}
&\pd{\uL}{t} = -  \frac{1}{2} C_3 \drat^{-3/2} \, \pd{}{x} \Proj (\uL)^2 - C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = \RHS[\uL]
%\qquad \text{for } x \in [-\pi,\pi]
\end{align}	
where we have represented the right-hand side by the operator $\RHS[\uL]$.
Our numerical scheme first separates the Hamiltonian into linear and nonlinear components. Then a pseudo-spectral scheme is used for spatial discretization with de-aliasing applied to the nonlinear term $\Proj (\uL)^2$ according to the standard 2/3-rule. That is, we first pad $\uhat_k$ with zeros for $\Lambda < \abs{k} \le 3\Lambda/2$, transform to physical space and square to obtain $\uL^2$ on a fine grid, then transform back to spectral space and truncate to obtain the Fourier coefficients of $\Proj (\uL)^2$
For simplicity, we denote these Fourier coefficients $\usqdah_k$. That is,
\begin{equation}
\Proj (\uL)^2 = \sum_{\abs{k} \le \Lambda} \usqdah_k e^{i k x}
\end{equation}
With this spectral discretization, the ODE system \eqref{TKdV2} can be written as
\begin{equation}
\label{TKdVDisc}
\td{}{t} \uhat_k =  -\frac{1}{2} C_3 \drat^{-3/2} \, ik \usqdah_k + C_2 \drat^{1/2} \, ik^{3} \uhat_k = \RHSh_k
\end{equation}
Due to the nonlinearity, this system mixes scales during evolution. We note the third-order linear term may become stiff for large $\Lambda$.

For time integration of \eqref{TKdVDisc}, we employ a 4th-order midpoint symplectic scheme \cite{mclachlan1993symplectic}. We introduce the spectral vectors $\mathbf{u}=\left\{ \hat{u}_{k}\right\}_{k=0}^{\Lambda}$ and $\mathbf{F}=\left\{ \RHSh_{k} \right\}_{k=0}^{\Lambda}$, and let $\mathbf{u}^{n}$ denote the solution at time $t_{n}$. The midpoint method has two intermediate stages and the auxiliary vectors $\mathbf{y}_{1},\mathbf{y}_{2}$:
\begin{align}
\label{SympSteps}
&\mathbf{y}_{1}-\mathbf{u}^{n} = w_{1}\Delta t \, \mathbf{F}
\left[ \frac{1}{2} \left(\mathbf{y}_{1}+\mathbf{u}^{n}\right) \right] ,\\
&\mathbf{y}_{2}-\mathbf{y}_{1} = w_{2}\Delta t \, \mathbf{F}
\left[ \frac{1}{2} \left(\mathbf{y}_{2}+\mathbf{y}_{1}\right) \right] ,\\
&\mathbf{u}^{n+1}-\mathbf{y}_{2} = w_{3}\Delta t \, \mathbf{F}
\left[ \frac{1}{2} \left(\mathbf{u}^{n+1}+\mathbf{y}_{2}\right) \right] ,
\end{align}
with the time increments $w_{1}=\left(2+2^{1/3}+2^{-1/3}\right)/3$,
$w_{2}=1-2w_{1}$, and $w_{3}=w_{1}$.

@@

\todo[inline]{Di, I do not understand this paragraph. Is the $F$ meant to be the same $F$ as above, i.e.~the right-hand-side of the TKdV equation \eqref{TKdV2}? Or is $F$ some arbitrary function of the state variable that we want to compute the expectation of?}
Then the PDFs can be counted from the solution $\uL(t)$, and the statistical expectations can be calculated by averaging along the trajectory
\begin{equation}
\left\langle F \left(u_{\Lambda}\right)\right\rangle _{\mu}=\frac{1}{T}\int_{t_{0}}^{t_{0}+T} F \left(u_{\Lambda}\left(t\right)\right)dt,
\end{equation}
given ergodicity in the solution with $\uL(t_0)$ already reaching the stationary state.

\todo[inline]{I am also a little confused about the paragraphs below. Eq.~\eqref{iterate} is not the same form of Eq.~(43)--(45) because in those equations it is not simply $\mathbf{z}$ inside $\mathbf{F}$, but instead an average of $\mathbf{z}$ and the previous value. Am I simply misinterpreting?}
\todo[inline]{Also, this is simply Picard iteration, correct? Perhaps we could simply say in words that we perform Picard iteration so that we don't have to worry about getting the equations correct in detail. What's your opinion?}

For the symplectic schemes, implicit iteration method is required to get the accurate solution. For any implicit problem of the form
\begin{equation}
\label{iterate}
\mathbf{z}-\mathbf{u}^{n}=\tau \mathbf{F \left(z\right)} = \tau\mathbf{Lz}+\tau\mathbf{N\left(z\right)},
\end{equation}
we split the operators into a linear and nonlinear components and only treat the nonlinear part on the right hand side for the previous stage in the iteration. Then we have the iterative scheme for the solution of implicit problem 
\begin{equation}
\left(1-\tau\mathbf{L}\right)\mathbf{z}^{\left[l+1\right]} = \mathbf{u}^{n}+\tau\mathbf{N}\left(\mathbf{z}^{\left[l\right]}\right).
\end{equation}
The linear operator $\mathbf{L}$ is the diagonal differential operator, thus it is easy to invert in these schemes. The termination criteria is determined by
\begin{equation}
\left\Vert \mathbf{z}^{\left[l+1\right]}-\mathbf{z}^{\left[l\right]}\right\Vert <\delta\;\mathrm{and}\;\left\Vert \mathbf{z}^{\left[l+1\right]}-\mathbf{z}^{\left[l\right]}\right\Vert \geq\left\Vert \mathbf{z}^{\left[l\right]}-\mathbf{z}^{\left[l-1\right]}\right\Vert .
\end{equation}
It is found that the iteration tolerance can be set as $\delta=1\times10^{-10}$.
Finally, the starting guess for the iterations can be found by quadratic extrapolation for the previous three stages $\left(t_{u},\mathbf{u}^{n}\right), \left(t_{u}-\tau,\mathbf{u}^{n-1}\right), \left(t_{u}-2\tau,\mathbf{u}^{n-2}\right)$. Therefore we can find the extrapolated value at $t=t_{n}+\beta\tau$ by
\begin{equation}
\mathbf{u}^{n+\beta}=\mathbf{Q}\left(t_{n}+\beta\tau\right)=\left[\frac{1}{2}\beta\left(\beta+3\right)+1\right]\mathbf{u}^{n}-\beta\left(\beta+2\right)\mathbf{u}^{n-1}+\frac{1}{2}\beta\left(\beta+1\right)\mathbf{u}^{n-2}.
\end{equation}
